---
title: "Lecture 7: Point Level Models - EDA and Simulation"
output:
  revealjs::revealjs_presentation:
    theme: night
    center: true
    transition: none
    incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(ggplot2)
library(dplyr)
library(mnormt)
library(SemiPar)
library(leaflet)
library(gstat)
library(sp)
```

# Class Intro

## Intro Questions 
- What are some basic guidelines for EDA for point referenced data?
- For Today:
    - More EDA
    - Data Simulation

## Spatial EDA Overview

1. Map of locations
2. Histogram or other distributional figure
3. 3D scatterplot
4. General Regression EDA
5. Variograms and variogram clouds
6. Anistopic diagnostics

# Scallops Data Example

## 1. Map of Locations
```{r}
data(scallop)
scallop <- scallop %>% mutate(log.catch = log(tot.catch + 1))
leaflet(scallop) %>% addTiles() %>% addCircles()
```

## 1. Map of Locations - Takeaways

_Goal_: Understand the sampling approach

- Is this a grid?
- Are there directions that have larger distances?
- How large is the spatial extent?

## 2. Histogram
```{r}
scallop %>% ggplot(aes(x=tot.catch)) + geom_histogram(bins=70) + ggtitle('histogram of scallop catch') + xlab('Number of Scallops')
```

## 2. Histogram
```{r}
scallop %>% ggplot(aes(x=log.catch)) + geom_histogram(bins=70) + ggtitle('histogram of log scallop catch (+1)') + xlab('Number of Scallops')
```

## 2. Histogram - Takeaways

_Goal:_ Identify a sampling distribution for the data

- Continuous or discrete data
- A linear model approach will be used for the response
- Spatial structure can also be included in generalized linear models
- Outliers are worth investigating, but a data point that does not fit the assumed model should not automatically be eliminated

## 3. 3D scatterplot
```{r}
scallop %>%  ggplot(aes(x=longitude, y = latitude)) + geom_point(aes(color=tot.catch)) + scale_colour_gradient2()
```

## 3. 3D scatterplot
```{r}
scallop %>% ggplot(aes(x=longitude, y = latitude)) + geom_point(aes(color=log.catch)) + scale_colour_gradient2()
```

## 3. 3D scatterplot - Takeaways

_Goal:_ Examine the spatial pattern of the response

- Again, this is the response not the residual
- Can also think about a contour plot (using some interpolation method)

## 4. General Regression EDA

- Assessing relationship between variable of interest and covariate information
- No covariates are present in the scallops data


## 5. Variograms and variogram clouds: Exercise
Explore the code below: what are the differences in the three variograms?
```{r, echo = T, eval = F}
coordinates(scallop) = ~longitude+latitude
class(scallop)
scallop.sp <- scallop
proj4string(scallop.sp) <- CRS("+proj=longlat +datum=WGS84")  
scallop.utm <- spTransform(scallop.sp, CRS("+proj=utm +zone=18 ellps=WGS84"))

plot(variogram(log.catch~1, scallop))
plot(variogram(log.catch~1, scallop.sp))
plot(variogram(log.catch~1, scallop.utm))
```

## 5. Variograms 
```{r,  eval = T}
coordinates(scallop) = ~longitude+latitude
scallop.sp <- scallop
proj4string(scallop.sp) <- CRS("+proj=longlat +datum=WGS84")  
scallop.utm <- spTransform(scallop.sp, CRS("+proj=utm +zone=18 ellps=WGS84"))

plot(variogram(log.catch~1, scallop.utm))
```

## 5. Variogram Cloud
```{r}
plot(variogram(log.catch~1, scallop.utm, cloud=T))
```

## 5. Variograms and variogram clouds: Takeaways

_Goal:_ Visually diagnose spatial structure

## 6. Anisotropy

_Goal:_ Determine if direction influencs spatial structure

# Anisotropy

## Directional Variogram
- All of the variograms we have looked at are isotropic

```{r}
plot(variogram(log.catch~1, scallop.sp, alpha = c(0, 45, 90, 135)))
```

## Separable Correlations Functions

- If the differences in spatial structure are directly related to two coordinate sets, we can create a stationary, anistropic covariance function

- Let $$cor(Y(\boldsymbol{s + h}), Y(\boldsymbol{s})) = \rho_1(h_y) \rho_2(h_x),$$
where $\rho_1()$ and $\rho_2()$ are proper correlation functions.
- A scaling factor, $\sigma^2,$ can be used to create covariance.

## Geometric Anistropy

- Another solution is the class of geometric anisotropic covariance functions with 
$$C(\boldsymbol{s} - \boldsymbol{s'}) = \sigma^2 \rho((\boldsymbol{s} - \boldsymbol{s'})^TB(\boldsymbol{s} - \boldsymbol{s'})),$$
where $B$ is positive definite matrix and $\rho$ is a valid correlation function
- $B$ is often referred to as a transformation matrix which rotates and scales the coordinates, such that the resulting transformation can be simplified to a distance.

## Sill, Nugget, and Range Anisotropy
- Recall the sill is defined as $\lim_{d \rightarrow \infty} \gamma(d)$
- Let $\boldsymbol{h}$ be an arbitrary separation vector, that can be normalized as $\frac{\boldsymbol{h}}{||\boldsymbol{h}||}$
- If $\lim_{a \rightarrow \infty} \gamma(a \times \frac{\boldsymbol{h}}{||\boldsymbol{h}||})$ depends on $\boldsymbol{h}$, this is referred to as sill anisotropy.
- Similarly the nugget and range can depend on $\boldsymbol{h}$ and give nugget anisotropy and range anisotropy 



# Model Simulation

## Simulating Spatial Process

- Soon we will look at fitting models for spatial point data
- Simulating data gives a deeper understanding of the model fitting process
- Simulate a mean-zero, isotropic spatial process with a spherical covariance function

## Simulating Spatial Process

1. Construct spatial locations
2. Calculate distances
3. Define covariance function and set parameters
4. Sample realization of the process
5. Visualize realization of spatial process

## 1. Construct Spatial Locations

```{r}
set.seed(01252019)
num.locations <- 50
coords <- data.frame(x = runif(num.locations), y = runif(num.locations))
coords %>% ggplot(aes(x=x,y=y)) + geom_point() + ggtitle('Hypothetical Sampling Locations') + xlim(0,1) + ylim(0,1)
```

## 2. Calculate Distances

```{r, echo = T}
dist.mat <- dist(coords, diag = T, upper = T) %>% as.matrix()
```

## 3. Define Covariance Function and Set Parameters
- Use spherical covariance with no nugget: 
$$C(d) = \begin{cases}
0 \text{ if } d \geq \frac{1}{\phi} \\
\sigma^2 \left[1 - \frac{3}{2}\phi d + \frac{1}{2} (\phi d)^3 \right] \; \; \text{ if } 0 \leq d \leq \frac{1}{\phi}
\end{cases}$$

```{r}
sigma.sq <- 1
phi <- 2
Sigma <- sigma.sq * (matrix(1,num.locations,num.locations) - 1.5 * phi * dist.mat + .5 * (phi * dist.mat)^3) 
Sigma[dist.mat > 1/phi] <- 0
```


## 4. Sample realization of the process
- This requires a distributional assumption, we will use the Gaussian distribution

```{r}
Y <- rmnorm(n=1, mean = 0, varcov = Sigma)
```

- What about the rest of the locations on the map?

## 5. Vizualize Spatial Process

```{r}
coords %>% mutate(Y = Y) %>% ggplot(aes(x=x,y=y)) + geom_point(aes(color=Y), size=2) + ggtitle(label = 'Simulated Spatial Process', subtitle = 'Spherical Covariance: sigma.sq = 1, phi = 2') +  xlim(0,1) + ylim(0,1) +   scale_colour_gradient2() + theme_dark()
```

## Extra: Gridded Spatial Process

```{r}
dim.grid <- 10
grid.coords <- data.frame(x.grid = rep(seq(.05, .95, length.out=dim.grid), dim.grid),
  y.grid = rep(seq(.05, .95, length.out = dim.grid), each = dim.grid))

dist.grid <- dist(grid.coords, diag = T, upper = T) %>% as.matrix()

sigma.sq <- 1
phi <- 2
Sigma <- sigma.sq * (matrix(1,dim.grid^2,dim.grid^2) - 1.5 * phi * dist.grid + .5 * (phi * dist.grid)^3) 
Sigma[dist.grid > 1/phi] <- 0

Y <- rmnorm(n=1, mean = 0, varcov = Sigma)

grid.coords %>% mutate(Y = Y) %>% ggplot(aes(x=x.grid,y=y.grid)) + geom_point(aes(color=Y), size=3) + ggtitle('Simulated Spatial Process', subtitle = 'Spherical Covariance: sigma.sq = 1, phi = 2') + xlim(0,1) + ylim(0,1) +   scale_colour_gradient2() + theme_dark()

```

## Extra: Gridded Spatial Process

```{r}
dim.grid <- 50
grid.coords <- data.frame(x.grid = rep(seq(.05, .95, length.out=dim.grid), dim.grid),
  y.grid = rep(seq(.05, .95, length.out = dim.grid), each = dim.grid))

dist.grid <- dist(grid.coords, diag = T, upper = T) %>% as.matrix()

sigma.sq <- 1
phi <- 2
Sigma <- sigma.sq * (matrix(1,dim.grid^2,dim.grid^2) - 1.5 * phi * dist.grid + .5 * (phi * dist.grid)^3) 
Sigma[dist.grid > 1/phi] <- 0

Y <- rmnorm(n=1, mean = 0, varcov = Sigma)

grid.coords %>% mutate(Y = Y) %>% ggplot(aes(x=x.grid,y=y.grid)) + geom_point(aes(color=Y), size=3) + ggtitle('Simulated Spatial Process', subtitle = 'Spherical Covariance: sigma.sq = 1, phi = 2') + xlim(0,1) + ylim(0,1) +   scale_colour_gradient2() + theme_dark()
```


## Simulated Spatial Process: Exercise

How does the spatial process change with:

- another draw with same parameters?
- a different value of $\phi$
- a different value of $\sigma^2$
- adding a nugget term, $\tau^2$

# Model Fitting

## Classical Model Fitting

- The classical approach to spatial prediction is rooted in the minimum the mean-squared error.
- This approach is often referred to as *Kriging* in honor of D.G. Krige a South African mining engineer.
- As a result of Krige's work (along with others), point-level spatial analysis and geostatistical analysis are used interchangeably.


## Additional Resources
- [Meuse Data Tutorial](https://cran.r-project.org/web/packages/gstat/vignettes/gstat.pdf)
- [Textbook Data Sets](https://www.counterpointstat.com/hierarchical-modeling-and-analysis-for-spatial-data.html/)
